---
title: "Data Science for Economists"
# subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
subtitle: "A Quick Primer on Auctions"
author: "Alex Marsh"
date: "University of North Carolina | [ECON 390](https://github.com/alexiom/ECON390)" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
name: toc

```{css, echo=FALSE}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center", 
  fig.height=4, #fig.width=6, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
library(fontawesome)
library(microbenchmark)
library(data.table)
library(ggplot2)
set.seed(123)
```

# Table of contents

1. [Auctions](#auctions)

---

class: inverse, center, middle
name: auctions

# Auctions

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Motivation

Auction Theory has been one of the successes of economics in the past 30-40 years. 

Since auctions are very common in the real world, economists who study empirical auctions have also cropped up to estimate the models the theorists have developed.

While we will not delve too deeply into auctions, I do need to give you just enough information for the problem set. 

---
# Auction Models

The way we model auctions is that bidders have some *valuation* $v_i$ that comes from some distribution $F_v$. 

The valuations could be iid or iid conidtional on observables or have correlated valuations. 

Bidders then know the rules of the auction and submit bids. They bid in such a way that maximizes their expected utility: $$\max_{b_i} v_i \text{Pr}(\text{win with } b_i) - E[p(b_i) | b_i]$$

I am keeping this vague because there are different auction formats and not all require paying the exact bid you submitted e.g. second price auctions.

Some auction formats: 
1. First-Price Sealed Bid
2. English (Ascending) 
3. Dutch
4. Second-Price Sealed Bid
5. All-Pay

---
# Auction Models (Cont.)

Once a bidder has her valuation, we think that she forms some optimal bidding strategy $b^*(v_i)$ that maps her valuation to a bid. We denote the observed bid $b_i$ and we say that $b_i = b^*(v_i)$ with equilibrium bidding.

In empirical auctions, the goal is to be able to recover $v_i$ or $F_v$ from a bunch of bidding data $(b_i)_{i=1}^{N}$, information about how the auction was ran, and assumptions about bidding behavior. 

For the problem set, the only auction format that we need to talk about are English Auctions. 

English Auctions start at 0 and bidders will continue to call out new prices until only one bidder remains. 
 - Sometimes there are minimum bid increments. 
 - In reality, there always is a minimum bid increment: the lowest denomination of currency e.g. a penny with US currency. 

```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
infile = list.files(pattern = '.html')
pagedown::chrome_print(input = infile, timeout = 100)
```